HealpixNet: Efficient spherical Convolutional Neural Network with Healpix sampling for cosmological applications

### 2018-07-18 scnn.tex
libsharp

The first approach, leveraging the continuous spherical convolution, has been proposed by~\citet{cohen2018spherical}. The goal of this contribution was to define a rotation-equivariant convolution on the sphere, i.e., a rotation of the input implies the same rotation in the output.
The resulting convolution is performed by (i) a spherical harmonic transform (SHT), i.e., a projection on the spherical harmonics, (ii) a  multiplication in the spectral domain, and (iii) an inverse SHT. Hence the computational cost of a convolution is essentially due to the two SHTs. A naive implementation of the SHT costs $\bO(N_{pix}^2)$ operations, where $N_{pix}$ is the number of pixels. There fortunately exists accelerated schemes for some sampling sets, see for example \citet{mohlenkamp1999fast} and \citet{rokhlin2006fast}.
The main advantage of this approach is that it provides a mathematically well defined rotation-equivariant network. Nevertheless, even with accelerated SHTs, the convolutions remain expensive, limiting the practical use of this approach.
For example with the HEALPix sampling, which was designed to have a fast SHT by being iso-latitude, the computational cost of the SHT is $\bO(N_{pix}^{3/2}) = \bO(N_{side}^3) = \bO(\ell_{max}^3)$, where $\ell_{max}$ is the largest order (angular frequency) \citep{gorski2005healpix}.\footnote{All pixels are placed on $N_{ring} = 4N_{side}-1 = \bO(\sqrt{N_{pix}})$ rings of constant latitude. Each ring has $\bO(\sqrt{N_{pix}})$ pixels. Thanks to this iso-latitude property, the SHT is computed using recurrence relations for Legendre polynomials on co-latitude and fast Fourier transforms (FFTs) on longitude. The computational cost is thus $\sqrt{N_{pix}}$ FFTs for a total cost of $\bO(N_{pix} \log(\sqrt{N_{pix}}))$, plus $\sqrt{N_{pix}}$ matrix multiplications of size $\sqrt{N_{pix}}$ for a total cost of $\bO(N_{pix}^{3/2})$.}
In comparison, filtering with our method scales as $\bO(N_{pix})$ (see \secref{efficient_convolution}). Moreover, \figref{filtering_speed} shows that a naive implementation of our method is more than ten times faster than the SHT implemented in HEALPix.
While a comparison is theoretically possible, the experiments of~\cite{cohen2018spherical} are done using the geodesic grid sampling set and do not provide any code for the HEALPix one. Hence, a practical comparison is not possible.
We also note that this approach cannot be easily accelerated when data lies on a part of the sphere only. That is an important use case in cosmology as measurements are often partial, i.e., full-sky maps are rare. One could still fill the unseen part of the sphere with zeros, and avoid to compute empty integrals (on the latitude or longitude). It is however not straightforward to identify those empty directions, and computations would still be wasted (consider for example a ring which mostly contains zeros but a few measurements). With graphs however, one only pays for the used pixels. The cost is some distortions due to border effects (see \figref{border_effects} and \ref{sec:border_effects}). Those can however be mitigated by padding with zeros a small area around the measurements.

>>>>

The first approach, leveraging the continuous spherical convolution, has been proposed by~\citet{cohen2018spherical}. The goal of this contribution was to define a rotation-equivariant convolution on the sphere, i.e., a rotation of the input implies the same rotation in the output.
The resulting convolution is performed by (i) a spherical harmonic transform (SHT), i.e., a projection on the spherical harmonics, (ii) a  multiplication in the spectral domain, and (iii) an inverse SHT. Hence the computational cost of a convolution is essentially due to the two SHTs. A naive implementation of the SHT costs $\bO(N_{pix}^2)$ operations, where $N_{pix}$ is the number of pixels. There fortunately exists accelerated schemes for some sampling sets, see for example \citet{mohlenkamp1999fast, rokhlin2006fast, reinecke2013libsharp}.
The main advantage of this approach is that it provides a mathematically well defined rotation-equivariant network. Nevertheless, even with accelerated SHTs, the convolutions remain expensive, limiting the practical use of this approach.
For example with the HEALPix sampling, which was designed to have a fast SHT by being iso-latitude, the computational cost of the SHT is $\bO(N_{pix}^{3/2}) = \bO(N_{side}^3) = \bO(\ell_{max}^3)$, where $\ell_{max}$ is the largest order (angular frequency) \citep{gorski2005healpix, reinecke2013libsharp}.\footnote{All pixels are placed on $N_{ring} = 4N_{side}-1 = \bO(\sqrt{N_{pix}})$ rings of constant latitude. Each ring has $\bO(\sqrt{N_{pix}})$ pixels. Thanks to this iso-latitude property, the SHT is computed using recurrence relations for Legendre polynomials on co-latitude and fast Fourier transforms (FFTs) on longitude. The computational cost is thus $\sqrt{N_{pix}}$ FFTs for a total cost of $\bO(N_{pix} \log(\sqrt{N_{pix}}))$, plus $\sqrt{N_{pix}}$ matrix multiplications of size $\sqrt{N_{pix}}$ for a total cost of $\bO(N_{pix}^{3/2})$.}
In comparison, filtering with our method scales as $\bO(N_{pix})$ (see \secref{efficient_convolution}). Moreover, \figref{filtering_speed} shows that a naive implementation of our method is more than ten times faster than the SHT implemented by the highly optimized \pkg{libsharp} library used by HEALPix \citep{reinecke2013libsharp}.
While a comparison is theoretically possible, the experiments of~\cite{cohen2018spherical} are done using the geodesic grid sampling set and do not provide any code for the HEALPix one. Hence, a practical comparison is not possible.
We also note that this approach cannot be easily accelerated when data lies on a part of the sphere only. That is an important use case in cosmology as measurements are often partial, i.e., full-sky maps are rare. One could still fill the unseen part of the sphere with zeros, and avoid to compute empty integrals (on the latitude or longitude). It is however not straightforward to identify those empty directions, and computations would still be wasted (consider for example a ring which mostly contains zeros but a few measurements). With graphs however, one only pays for the used pixels. The cost is some distortions due to border effects (see \figref{border_effects} and \ref{sec:border_effects}). Those can however be mitigated by padding with zeros a small area around the measurements.
